---
title: "pomp data"
author: "Andrew Tredennick"
date: "4/5/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) 
library(pomp)
```


**`pomp` wants time-indexed case reports when using `accumvars` that reset to 0 at each observation timestep.**

That's the punchline; let's prove it.

## Evidence from examples

The codes for King's `dacca`, `ebola`, and `sir` examples all have time-indexed case reports (new cases or some index thereof) as the data.
None of these use cumulative case counts as the data.
Likewise, these all use the `accumvar` definition for the variables passed to the likelihood evaluator.
 
Here's the `dmeasure` snippet for the `sir` example explored below:
```{r dm}
dmeasure = Csnippet(
  " double mean, sd;
    double f;
    mean = cases*rho;
    sd = sqrt(cases*rho*(1-rho));
    if (reports > 0) {
      f = pnorm(reports+0.5,mean,sd,1,0)-pnorm(reports-0.5,mean,sd,1,0);
    } else {
      f = pnorm(reports+0.5,mean,sd,1,0);
    }
    lik = (give_log) ? log(f) : f;"
  )
```

And here's the `accumvar` setting for the simulated `cases` variable that is evaluated relative to the `reports` data in the likelihood function.

```{r acc}
accumvars = c("cases")
```

As we'll see below, both `cases` (sims) and `reports` (data) need to both be new cases, not cumulative cases.


## Evidence from quick simulation

Here I use the `pomp::sir` model and simulate some data.
First, I use `pfilter` to compute the likelihood of the model fit to time-indexed reports of new cases.
Second, I use the same procedure to the likelihood of the model fit to cumulative reports.

### Model set up
Note that the data supplied with the pomp object are new reports (they go up and down)
```{r modset}
sir_mod <- sir()  # load the pomp object

plot(t(sir_mod@data))  # data that is shipped with pomp
```

### Simulate data
```{r simdat}
# Simulate a data set
sim_data <- simulate(sir_mod, nsim = 1, format="data.frame", 
                     include.data = FALSE)

sir_mod@data[1, ] <- sim_data$reports  # replace data

coef(sir_mod)  # print the coefficients
```

### Fit to new reports
```{r fit}
pf <- pfilter(sir_mod, Np = 2000)
plot(pf)
```

## Fit to cumulative reports
Note that there are many, many filtering failures; that the "reports" (data) and cases (sims) do not match up; and that the likelihood cannot even be evaluated.

```{r fit2}
sir_mod@data[1, ] <- cumsum(sim_data$reports)  # replace data with cumulatives
pf2 <- pfilter(sir_mod, Np = 2000)
plot(pf2)
```

## Conclusion

We need to use daily reports of new cases as the data in our fitting routines.
This means using ```diff(covid_ga_data$cases)``` as the data.